---
title: "Lab 5: Visualizing Coronavirus Data"
author: "Name: Sal Figueroa"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
header-includes:
#These packages are not necessary to complete the assignment and or only used to provide an example. 
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
  - \usepackage{ggrepel}
---

```{r setup, include = FALSE}
#### Load necessary library ####
packages <- c("knitr", "kableExtra", "magrittr", "readr", "geosphere")

install_me <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(install_me)) install.packages(install_me)

library(knitr)
library(magrittr)
library(readr)
library(geosphere)
library(ggplot2)
library(treemapify)
library(tidyverse)
library(ggfittext)
library(scales)
library(dplyr)
library(data.table)
library(ggrepel)
library(scales)
library(stringr)

knitr::opts_chunk$set(echo=TRUE)

```


```{r Data setup, include = FALSE}
#time_series_covid19_confirmed_US.csv 1154 columns, 289 rows.
Confirmed_Global <- as.data.frame(read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv"))

#Serve as data frame length and width variables if needed
rowmax <- nrow(Confirmed_Global[,]) #limit Qty of Rows 289
colmax <- ncol(Confirmed_Global[,]) #limit Qty of Rows 1147

#loads column headers in data.frame
NameCol <- as.data.frame(colnames(Confirmed_Global))

#Var holds the length of the NameCol data.frame
MAXNameCol <- nrow(NameCol)

#Name of the final column name
FinalColName <- NameCol[MAXNameCol,] 


#Loads the URL of the .RData file in the GitHub repository
GitHuburl <- "https://github.com/Figgs0bit/CSIT165-ModData/raw/refs/heads/master/Module-13/human_proteins.RData"
#names the download for the *.Rdata file.
RdataFile <- "human_proteins.RData"

#Download the file into the working directory, Note: mode-WB needed to download binary .rdata file.
download.file(GitHuburl, RdataFile, mode = "wb")
#Load the .RData file from working directory into the R environment
load(RdataFile)

```



## Github Repository

*Holds all related files*

[github: Figgs0bit-Lab4 (https://github.com/Figgs0bit/CSIT165-Lab5.git)](https://github.com/Figgs0bit/CSIT165-Lab5)


## Required data sets

*This lab represents data downloaded on 04/xx/2025*

[Human Proteins Data Set ()](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)


### *1. 2019 Novel Coronavirus COVID-19 (2019-nCoV) Global Confirmations.*

*This data set is operated by the John Hopkins University Center for Systems Science and Engineering (JHU CSSE). Data set includes a daily time series     CSV summary table confirmed cases of COVID-19. Lat and Long refer to coordinate references for the data field. Date fields are stored in MM/DD/YYYY        format.*


[2019 Novel Coronavirus COVID-19 (2019-nCoV) Global         Confirmations](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)


### *2. Human Proteins Data Set*

*This data set is a tibble created from parsing Homo sapiens protein fasta files curated by the Genome Reference Consortium as part of the Human Genome   Project. The original protein fasta file can be found in NCBI, here. Data consists of two columns, Gene and Protein.Sequence. Gene represents every gene   product, or protein, made by humans. Protein.Sequence represents the primary amino acid structure of its correspondent gene. Each amino acid is            represented as a single capital letter and the sequence of letters is unique to each gene.*


## Instructions

*Before beginning your objectives in your final document, please state which day you downloaded the data sets on for analysis. The objectives for this lab will cumulatively cover many subjects discussed in this course and will also contain an objective for manipulating strings.*

*The surgeon general for the United States recently created a new data science initiative, CSIT-165, that uses data science to characterize pandemic diseases. CSIT-165 disseminates data driven analyses to state governors. You are a data scientist for CSIT-165 and it is up to you and you alone to manipulate and visualize COVID-19 data for disease control.*


## Objective 1

*Create a scatter plot for counts per day of the the top five confirmed countries. For this objective, please use dplyr and tidyr to manipulate data and ggplot2 to create the visualization. Scatter plot must have specified colors, a non-standard theme for display, and custom a customized titles, axis labels, and legend labels.*


```{r ob1}

#Ensure FinalColName is treated as a string
FinalColName <- as.character(FinalColName)

#Load Confirmed_Global into temp variable
DF <- Confirmed_Global

#Reorganize the data frame with the column specified in FinalColName in descending order
Decrease_DF <- DF[order(DF[[FinalColName]], decreasing = TRUE),]


#Creates data frame with the top 5 Total unaccounted cases.
#Here the "head" command trims away any row past the 5th row. 
top5_DF <- head(Decrease_DF, 5)

#Creates a vector full of Dates reflective of the data frame dates. 
Date_Days <- as.Date(seq(as.Date("2020/1/22"), as.Date("2023/3/9"), "days"))


#Creates Vectors for the Top 5 Countries Data
#Rpw 1: Country
rowOne <- as.vector(t(top5_DF[1,]))
rowOne <- as.numeric(rowOne[5:MAXNameCol])
#Rpw 2: Country
rowTwo <- as.vector(t(top5_DF[2,]))
rowTwo <- as.numeric(rowTwo[5:MAXNameCol])
#Rpw 3: Country
rowThree <- as.vector(t(top5_DF[3,]))
rowThree <- as.numeric(rowThree[5:MAXNameCol])
#Rpw 4: Country
rowFour <- as.vector(t(top5_DF[4,]))
rowFour <- as.numeric(rowFour[5:MAXNameCol])
#Rpw 5: Country
rowFive <- as.vector(t(top5_DF[5,]))
rowFive <- as.numeric(rowFive[5:MAXNameCol])

#Creates Data.frame with the dates and the top five countries.
#The data has now been formatted as in an Inverse matrix format
T_top5_DF <- data.frame(Date_Days, rowOne, rowTwo, rowThree, rowFour, rowFive)

#Renames the column names
colnames(T_top5_DF) <- c("Date", top5_DF[1,2], top5_DF[2,2], top5_DF[3,2], top5_DF[4,2], top5_DF[5,2])

#Converts data.frame to long format: assumes T_top5_DF has Date + 5 country columns
#-Date excludes the date here. Serves to plot multiple country data into one plot window. 
T_top5_long <- T_top5_DF %>%
  pivot_longer(-Date, names_to = "Country", values_to = "Cases")



# Plot all 5 countries in one panel
ggplot(T_top5_long, aes(x = Date, y = Cases, color = Country)) + #Loads each of the top 5 couties data via color = country

  geom_line(size = 0.125) + #Sets thickness of line

  #Plot labels, title, x & y axis
  labs(title = "COVID Data for Top 5 Countries", x = "Date: (Year/Day/Month)", y = "Confrimed Global Cases") + 

  theme_minimal(base_size = 8) + #Creates simple plot theme w/ a size 8 font size

  theme(axis.text.x = element_text(angle = 90, hjust = 1), #rotates x-axis text and hjust aligns the text to the right
        axis.text.y = element_text(angle = 315, hjust = 1), #rotates y-axis text and hjust aligns the text to the right
        axis.title.y = element_text(margin = margin(r = 12)), #adds margin to y axis
        panel.grid.major = element_line(size = 0.2), #major grid lines
        panel.grid.minor = element_line(size = 0.0625) #minor grid lines
       )

```

## Objective 2

*Understanding how COVID-19 enters human cells requires that we have a better understanding of human proteins. It has been shown that COVID-19 is able to enter cells by binding to Angiotensin-Converting Enzyme 2 receptors in the heart, lungs, and intestines. Angiotensin-Converting Enzyme 2 is used by the body to regulate blood pressure and inflammation.*

*1.	Show how many different isoforms of Angiotensin-Converting Enzyme 2 humans make using str_detect and regular expressions.*
*Use the pattern “Angiotensin-Converting Enzyme 2 isoform” with regular expressions to include all variations irrespective of first letter capitalization.*


```{r ob2.1}
#Renames human_protiens data frame
proteins <- human_proteins
# Check column names
#str(proteins)

ace2_isoforms <- proteins %>%
  filter(str_detect(Gene, regex("Angiotensin-Converting Enzyme 2 isoform", ignore_case = TRUE)))

head(ace2_isoforms)
```

*2.	Show the amino acid sequence between the 27th and 63rd amino acid sequence for each isoform using str_sub.*

```{r ob2.2}
#Use str_sub to extract positions 27 to 63 from Protein.Sequence
ace2_isoforms <- ace2_isoforms %>%
  mutate(seq_27_63 = str_sub(Protein.Sequence, 27, 63))

# Display the extracted subsequence
ace2_isoforms %>% select(Gene, seq_27_63)
```

*3.	Print a statement using cat with paste or sprintf of the first thirty amino acids for each isoform with a new line after each isoform and sequence listing. cat is necessary to use in combination with paste or sprintf to output concatenation with a new line.*

```{r ob2.3}
cat("First 30 amino acids of each isoform:\n\n")

#for loop searches for string mathches
for (i in 1:nrow(ace2_isoforms)) {
  gene_name <- ace2_isoforms$Gene[i]
  seq30 <- str_sub(ace2_isoforms$Protein.Sequence[i], 1, 30)
  cat(sprintf("%s: %s\n\n", gene_name, seq30))
}
```


